---
title: "Tree Biomass Estimation with PSP Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tree Biomass Estimation with PSP Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

## Overview

Aboveground tree biomass is used in carbon accounting, yield monitoring, and
nutrient cycling studies. In PSP programs, every remeasurement cycle produces
a fresh set of DBH observations; heights are measured for a subset of trees.
This vignette demonstrates how to use `BCallometryR` to estimate biomass from
PSP-style data, covering three common scenarios:

1. **DBH-only** — heights not available (most common in large PSP archives)
2. **DBH + height** — for trees where height was measured
3. **Component breakdown** — wood, bark, branches, foliage separately

The equations used are from Lambert et al. (2005) and Ung et al. (2008) —
Canadian national allometric equations fitted to destructively sampled trees.

---

## 1. Load PSP Data

The built-in `psp_trees` dataset contains 20 plots across four BEC zones
(SBS, ICH, CWH, IDF) with five BC species recorded as raw PSP/VRI inventory
codes (`AT`, `FDI`, `HW`, `PLI`, `SW`). Approximately 25% of trees have
measured heights; the rest have DBH only, which is the typical situation in a
BC PSP archive. `bc_species_to_biomass_name()` crosses the raw codes to the
common names expected by `biomass_tree()`.

```{r load-data}
library(CanWestAllometryCFS)

sim_trees <- psp_trees

# Crosswalk PSP codes to biomass common names
sim_trees$SPECIES_NAME <- bc_species_to_biomass_name(sim_trees$SPECIES,
                                                     sim_trees$BEC_ZONE)

cat(sprintf("Plots: %d   Trees: %d   Heights measured: %d (%.0f%%)\n",
            length(unique(sim_trees$SITE_IDENTIFIER)), nrow(sim_trees),
            sum(!is.na(sim_trees$HEIGHT)),
            100 * mean(!is.na(sim_trees$HEIGHT))))
```

---

## 2. DBH-Only Biomass

When tree heights are not available, the DBH-only equations provide a
reasonable biomass estimate. This is the most common case for historic PSP
data where full-height measurements were not collected.

```{r biomass-dbh-only}
sim_trees$BIOMASS_DBH <- biomass_tree(
  species = sim_trees$SPECIES_NAME,
  dbh     = sim_trees$DBH
)

# Quick summary
sp_summary <- do.call(rbind, lapply(sort(unique(sim_trees$SPECIES)), function(sp) {
  sub <- sim_trees[sim_trees$SPECIES == sp, ]
  data.frame(
    species   = sp,
    n         = nrow(sub),
    dbh_mean  = round(mean(sub$DBH), 1),
    bio_mean  = round(mean(sub$BIOMASS_DBH, na.rm = TRUE), 1),
    bio_range = paste0(round(range(sub$BIOMASS_DBH, na.rm = TRUE), 1), collapse = "–")
  )
}))
print(sp_summary, row.names = FALSE)
```

---

## 3. DBH + Height Biomass

For trees where height was measured, the two-variable equations generally
reduce prediction error, especially for large trees and species with strong
height allometry.

When `height` is supplied as a vector, `biomass_tree()` selects the equation
per tree: trees with a non-`NA` height use the DBH + height equations;
trees with `NA` height automatically fall back to the DBH-only equations.
This means no two-pass split is needed — the entire dataset can be passed in
one call.

```{r biomass-with-height}
# Single call: height-included where HEIGHT is available, DBH-only otherwise
sim_trees$BIOMASS_HT <- biomass_tree(
  species = sim_trees$SPECIES_NAME,
  dbh     = sim_trees$DBH,
  height  = sim_trees$HEIGHT
)

# Compare DBH-only vs DBH+height for trees where we have both
has_ht   <- !is.na(sim_trees$HEIGHT)
both     <- sim_trees[has_ht, ]
diff_pct <- 100 * (both$BIOMASS_HT - both$BIOMASS_DBH) / both$BIOMASS_DBH

cat(sprintf("Trees with both methods: %d\n", nrow(both)))
cat(sprintf("Mean difference (HT - DBH): %.1f%%\n", mean(diff_pct, na.rm = TRUE)))
cat(sprintf("SD of difference: %.1f%%\n",            sd(diff_pct,   na.rm = TRUE)))
```

```{r plot-comparison, fig.cap = "DBH+height vs DBH-only biomass for measured trees. The dashed line is 1:1."}
plot(BIOMASS_DBH ~ DBH, data = both,
     col  = "#00000040", pch = 16, cex = 0.7,
     xlab = "DBH (cm)", ylab = "Biomass (kg)",
     main = "Biomass estimates by method",
     ylim = range(c(both$BIOMASS_DBH, both$BIOMASS_HT), na.rm = TRUE))

points(BIOMASS_HT ~ DBH, data = both,
       col = "#d6604d80", pch = 16, cex = 0.7)

legend("topleft",
       legend = c("DBH only", "DBH + height"),
       col    = c("#000000", "#d6604d"),
       pch    = 16, bty = "n")
```

---

## 4. Component Breakdown

`biomass_components()` returns wood, bark, branches, and foliage separately.
This is useful when carbon fractions differ by component, or when only
certain biomass pools are relevant to a study.

```{r components}
# Component breakdown for a subset of trees (DBH-only for simplicity)
example_trees <- sim_trees[c(1, 15, 30, 55), ]

comp <- biomass_components(
  species = example_trees$SPECIES_NAME,
  dbh     = example_trees$DBH
)

print(cbind(
  species = example_trees$SPECIES_NAME,
  dbh     = example_trees$DBH,
  round(comp, 2)
), row.names = FALSE)
```

```{r plot-components, fig.cap = "Component biomass for four example trees (DBH-only equations)."}
bp <- barplot(
  t(as.matrix(comp[, c("wood", "bark", "branches", "foliage")])),
  names.arg = paste0(example_trees$SPECIES_NAME, "\n",
                     example_trees$DBH, " cm"),
  col  = c("#4d9221", "#a1d76a", "#e9f5e1", "#fdae61"),
  legend.text = c("wood", "bark", "branches", "foliage"),
  args.legend = list(x = "topleft", bty = "n", cex = 0.8),
  ylab = "Biomass (kg)",
  main = "Component biomass by tree",
  cex.names = 0.75
)
```

The wood component dominates for large conifer stems; branches and foliage
matter more for the foliage-rich smaller diameter classes and deciduous species.

---

## 5. Lambert 2005 vs Ung 2008

Both source papers provide national equations. Ung et al. (2008) re-fitted
some species including BC data; for species covered by both sources the
differences are generally small (< 5%).

```{r compare-sources}
# Compare for lodgepole pine, white spruce, trembling aspen across a DBH range
dbh_seq <- seq(10, 55, by = 5)

cmp <- lapply(c("lodgepole pine", "white spruce", "trembling aspen"), function(sp) {
  bl <- biomass_tree(sp, dbh_seq, paper_source = "Lambert2005")
  bu <- biomass_tree(sp, dbh_seq, paper_source = "Ung2008")
  data.frame(
    species       = sp,
    dbh           = dbh_seq,
    Lambert2005   = round(bl, 2),
    Ung2008       = round(bu, 2),
    diff_pct      = round(100 * (bu - bl) / bl, 1)
  )
})
cmp_df <- do.call(rbind, cmp)

# Show just the boundary DBH classes
print(cmp_df[cmp_df$dbh %in% c(10, 30, 55), ], row.names = FALSE)
```

---

## 6. Plot-Level Aggregation

PSP analysis typically aggregates to the plot level, often expressing biomass
as Mg/ha. Here we assume each plot is a fixed-area subplot (400 m², 0.04 ha),
which gives an expansion factor of 25.

```{r plot-level}
PLOT_AREA_HA <- 0.04          # 20 m × 20 m subplot
EXP_FACTOR   <- 1 / PLOT_AREA_HA  # trees/ha expansion factor

# Use DBH-only biomass for all trees (universal coverage)
plot_biomass <- do.call(rbind, lapply(unique(sim_trees$SITE_IDENTIFIER), function(pid) {
  sub   <- sim_trees[sim_trees$SITE_IDENTIFIER == pid, ]
  total <- sum(sub$BIOMASS_DBH, na.rm = TRUE) * EXP_FACTOR / 1000  # Mg/ha
  data.frame(
    SITE_IDENTIFIER = pid,
    BEC_ZONE   = sub$BEC_ZONE[1],
    N_TREES    = nrow(sub),
    BIOMASS_MG_HA = round(total, 1)
  )
}))

print(plot_biomass[order(plot_biomass$BIOMASS_MG_HA, decreasing = TRUE), ],
      row.names = FALSE)
cat(sprintf("\nMean plot biomass: %.1f Mg/ha  (SD: %.1f)\n",
            mean(plot_biomass$BIOMASS_MG_HA),
            sd(plot_biomass$BIOMASS_MG_HA)))
```

---

## Summary

| Goal | Function |
|------|----------|
| Total aboveground biomass (kg) | `biomass_tree(species, dbh)` |
| Total biomass with height | `biomass_tree(species, dbh, height)` |
| Component breakdown | `biomass_components(species, dbh, height)` |
| Use Ung 2008 equations | `biomass_tree(..., paper_source = "Ung2008")` |
| List available species | `sort(unique(biomass_coefs$species))` |
| View citations | `biomass_citations()` |

```{r citations}
biomass_citations()
```
