---
title: "Complete PSP Workflow: Species Crosswalk, H-D, Biomass, and Volume"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete PSP Workflow: Species Crosswalk, H-D, Biomass, and Volume}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

## Overview

This vignette demonstrates the end-to-end workflow for a simplified PSP
dataset with tree-level measurements.

The steps are:

1. **Species code crosswalk** — translate raw species codes (e.g. `"PLI"`,
   `"FDI"`, `"SW"`) to the group-level SP0 codes and species common names used
   by `tree_volume()` and `biomass_tree()`, respectively
2. **H-D model fitting** — where there are enough plots per species group and enough trees, fit a mixed-effects
   model; otherwise fit a fixed-effects model
3. **Height imputation** — fill missing heights using a localised
   mixed-effects prediction where applicable, or a population-average
   prediction where a mixed-effects model is not available
4. **Biomass estimation** — use the national-level biomass equations to
   estimate total aboveground biomass and a component breakdown
5. **Volume estimation** — calculate whole-stem volume (WSV) and gross
   merchantable volume (MER)

---

## 1. PSP data

Load the simplified PSP data included with the package:

```{r load-data}
library(CanWestAllometryCFS)

trees <- psp_trees

cat(sprintf(
  "%d plots | %d trees | live: %d | dead: %d | heights measured: %d (%.0f%%) | btop: %d\n",
  length(unique(trees$SITE_IDENTIFIER)), nrow(trees),
  sum(trees$LV_D == "L"), sum(trees$LV_D == "D"),
  sum(!is.na(trees$HEIGHT)), 100 * mean(!is.na(trees$HEIGHT)),
  sum(trees$BTOP)
))
```

In practice, any tree-level PSP data can be used. At minimum, the data should
have the following columns:

| Column | Description |
|--------|--------------|
| `SITE_IDENTIFIER` | Plot identifier |
| `BEC_ZONE` | BEC zone code (used for species correction and taper equation) |
| `SPECIES` | Raw BC inventory code, e.g. `"PLI"`, `"FDI"`, `"SW"`, `"HW"` |
| `DBH` | Diameter at breast height (cm) |
| `HEIGHT` | Total height (m); `NA` for unmeasured, dead, or broken-top trees |
| `LV_D` | Tree status: `"L"` = live, `"D"` = dead (standing snag) |
| `BTOP` | Logical; `TRUE` if the tree has a broken top |

---

## 2. Species code crosswalk

In BC, field-entered species codes are typically stored as 2–3 character codes.
The H-D modelling, volume, and biomass functions require either group-level
codes or species common names. The species code crosswalks are handled by
`species_correction()`, `bc_species_to_sp0()`, `bc_species_to_sp_type()`,
and `bc_species_to_biomass_name()`.

`species_correction()` is a necessary first step: it (1) truncates 3-letter
codes to their 2-letter form (e.g. `PLI` → `PL`, `FDI` → `FD`), and (2) uses
BEC zone to resolve ambiguous generic codes — for example, `"S"` becomes `"SW"`
in interior zones or `"SS"` in coastal zones, and `"B"` becomes `"BL"` or
`"BA"` depending on zone. Once the codes are unambiguous 2-letter codes,
`bc_species_to_sp0()` maps them to the group-level SP0 code (e.g. `"F"`,
`"H"`, `"S"`) and `bc_species_to_sp_type()` maps them to a broad
conifer/deciduous type (`"C"` or `"D"`). SP0 is the primary grouping used
by the H-D and volume functions; SP_TYPE is the fallback used by
`fit_hd_models_by_group()` when a species has too few measured trees to
fit its own SP0-level model.

The species code crosswalk pipeline is:

```
raw PSP code
  → species_correction(bec_zone)   # truncate + BEC-disambiguate (e.g. PLI→PL, S→SW/SS)
  → bc_species_to_sp0()            # → SP0  (e.g. "PL", "S", "F", "H")
  → bc_species_to_sp_type()        # → SP_TYPE  ("C" conifer or "D" deciduous)
  → bc_species_to_biomass_name()   # → common name  (e.g. "lodgepole pine")
```

```{r crosswalk}
trees$SPECIES_CORR    <- species_correction(trees$SPECIES, trees$BEC_ZONE)
trees$SPECIES_SP0     <- bc_species_to_sp0(trees$SPECIES_CORR)
trees$SPECIES_SP_TYPE <- bc_species_to_sp_type(trees$SPECIES_CORR)  # "C" or "D"
trees$SPECIES_NAME    <- bc_species_to_biomass_name(trees$SPECIES, trees$BEC_ZONE)

# Show unique crosswalk mappings
xwalk <- unique(trees[, c("SPECIES", "BEC_ZONE", "SPECIES_CORR",
                           "SPECIES_SP0", "SPECIES_SP_TYPE", "SPECIES_NAME")])
print(xwalk[order(xwalk$BEC_ZONE, xwalk$SPECIES), ], row.names = FALSE)
```

---

## 3. H-D model fitting

### Recommended pre-filtering

Before fitting H-D models, the training data should be restricted to
observations that represent a reliable measured height–DBH relationship.
The following filters are recommended and mirror those used in the BC MoF PSP and non-PSP compilation routines:

| Filter | Rationale |
|--------|-----------|
| Height measured (`!is.na(HEIGHT)`) | Only trees with a recorded height can be used to fit the H-D relationship |
| Broken-top trees excluded | no valid height–DBH observation to train on |
| Alive trees only (`LV_D == "L"`) | Dead trees are not used in H-D modelling |
| Implausible heights excluded | Trees with abnormal height–DBH ratios should not be used; what constitutes "abnormal" is at the user's discretion (a useful starting point: flag trees where `HEIGHT / DBH ≥ 3`) |
| Implausible DBH excluded | Trees with clearly erroneous DBH values (e.g. a dramatic unexplained drop between visits) should be excluded; again at the user's discretion |
| Height > 1.3 m and DBH > 0 | Removes seedlings and invalid records |

The recommended H-D model fitting strategy follows a two-level approach:

1. **SP0 level** — fit one model per SP0-level group (e.g. all spruce)
2. **SP_TYPE fallback** — if an SP0 group has too few measured trees or fails
   to converge, the broader SP_TYPE group (conifer `"C"` or deciduous `"D"`)
   model is used instead

`fit_hd_models_by_group()` will, by default, fit an H-D model by SP0, falling
back to the SP_TYPE group when there are too few plots or too few trees per
SP0 group.

For training data with repeated measurements (i.e. multi-year PSP data), set
`site_col = "SITE_IDENTIFIER"` (plot level) and `tree_col = "unitreeid"`
(individual tree level), where the plot-level effect absorbs between-plot
productivity differences and the tree-level effect accounts for repeated
measurements of the same tree across years.

The training data may include both plots that have multiple measurement
cycles and plots that have only one measurement cycle.

The following decision tree shows the internal logic used by
`fit_hd_models_by_group()`. Users may adjust the tree and plot thresholds
through the `min_n` and `min_plots` arguments.

> **Note on thresholds:** The defaults (`min_n = 10`, `min_plots = 5`) are
> set for small study-area datasets. The BC MoF compilation routine fits
> species- and SP0-level models only when **≥ 1 000 observations** are
> available from its province-wide PSP + non-PSP pool. Users working with
> province-wide data may want to set `min_n = 1000` to match that standard.

The decision tree for each SP0 group:

```
For each SP0 group (all measured trees pooled across years)
│
│  n_obs  = trees with BOTH height and DBH recorded for this SP0 group
│  n_plots = number of distinct plots containing at least one such tree
│
├─ n_obs < 10  OR  fitting fails  ──────────────►  no SP0 model
│                                                   → SP_TYPE fallback below
│
└─ n_obs ≥ 10
   │
   ├─ n_plots ≥ 5
   │  ├─ mixed-effects converges  ────────────────►  mixed-effects  ✓
   │  └─ mixed-effects fails  →  nls succeeds  ───►  fixed-effects  ✓
   │
   └─ n_plots < 5
      └─ nls succeeds  ──────────────────────►  fixed-effects  ✓

SP_TYPE fallback (for SP0 groups with no model):
└─ Pool all conifers ("C") or all deciduous ("D")
   └─ apply the same mixed-effects → fixed-effects decision as SP0 level above
```

```{r hd-fitting}
# Training data: alive trees with a measured height, excluding broken-top trees.
# For real PSP data also screen for implausible H/DBH ratios and erroneous DBH values.
measured  <- trees[!is.na(trees$HEIGHT) & !trees$BTOP & trees$LV_D == "L", ]
hd_result <- fit_hd_models_by_group(measured)
print(hd_result$summary, row.names = FALSE)

# SP0 list used by the biomass and volume summary chunks below
sp0_list <- sort(unique(trees$SPECIES_SP0[!is.na(trees$SPECIES_SP0)]))
```

---

## 4. Height imputation

`ht_impute()` fills missing heights in the full data frame using the models
from Section 3.  It applies the SP0-level model first, falling back to the
SP_TYPE model for any tree whose species has no SP0 model.  Three columns are
added to `trees`:

- `HT_PROJ` — filled height (m); measured heights are carried through unchanged
- `HD_SOURCE` — source flag (e.g. `"measured"`, `"naslund_mixed-effects_sp0"`)
- `HT_FLAG` — `"btop"` for broken-top trees, `NA` otherwise

For users seeking to replicate BC MoF's compilation routine, set
`impute_btop = TRUE`. This will estimate a total height from DBH for
broken-top trees using the H-D models fitted in Section 3. Estimated heights
are listed in `HT_PROJ`. The column `HT_FLAG` is used to show estimated
heights on trees with a broken top.

If break height and diameter at the break are available, `ht_from_btop()` can
be used instead to recover total height via a taper equation, which is more
precise than the H-D model approach. Set `impute_btop = FALSE` (the default)
if you intend to handle broken-top trees separately this way.

```{r height-imputation}
trees <- ht_impute(trees, hd_result, impute_btop = TRUE)

cat(sprintf("Heights filled: %d  |  Measured: %d  |  Still missing: %d\n",
            sum(!is.na(trees$HT_PROJ) & is.na(trees$HEIGHT)),
            sum(!is.na(trees$HEIGHT)),
            sum(is.na(trees$HT_PROJ))))

print(table(HD_SOURCE = trees$HD_SOURCE, useNA = "ifany"))
print(table(HT_FLAG   = trees$HT_FLAG,   useNA = "ifany"))
```

---

## 5. Biomass estimation

`biomass_tree()` uses height-included equations when `height` is supplied and
DBH-only equations when it is not.  Because `impute_btop = TRUE` was used
above, every tree has an `HT_PROJ` and biomass can be estimated in a single
pass.  `biomass_components()` returns wood, bark, branches, and foliage
separately if component-level carbon fractions are needed.

```{r biomass}
# biomass_tree() selects the equation per tree: height-included where
# HT_PROJ is not NA, DBH-only where it is. No two-pass split needed.
has_sp <- !is.na(trees$SPECIES_NAME)

trees$BIOMASS_KG <- NA_real_
trees$BIOMASS_KG[has_sp] <- biomass_tree(
  trees$SPECIES_NAME[has_sp], trees$DBH[has_sp],
  height = trees$HT_PROJ[has_sp])

comp <- biomass_components(trees$SPECIES_NAME[has_sp],
                           trees$DBH[has_sp],
                           height = trees$HT_PROJ[has_sp])

for (fld in c("wood", "bark", "branches", "foliage")) {
  col <- paste0("BIO_", toupper(fld), "_KG")
  trees[[col]]          <- NA_real_
  trees[[col]][has_sp]  <- comp[[fld]]
}

bio_summary <- do.call(rbind, lapply(sp0_list, function(sp0) {
  sub <- trees[!is.na(trees$SPECIES_SP0) & trees$SPECIES_SP0 == sp0 &
               !is.na(trees$BIOMASS_KG), ]
  data.frame(SP0      = sp0,
             n        = nrow(sub),
             mean_kg  = round(mean(sub$BIOMASS_KG), 1),
             sd_kg    = round(sd(sub$BIOMASS_KG),   1))
}))
print(bio_summary, row.names = FALSE)
```

---

## 6. Volume estimation

`tree_volume()` requires a height for every tree; trees without a filled
height are excluded (flagged by `has_ht`). Both WSV and gross merchantable
volume are computed in a single pass each.

```{r volume}
has_ht <- !is.na(trees$HT_PROJ) & !is.na(trees$SPECIES_SP0)

trees$WSV_M3 <- NA_real_
trees$MER_M3 <- NA_real_

trees$WSV_M3[has_ht] <- tree_volume(
  bec_zone = trees$BEC_ZONE[has_ht],
  species  = trees$SPECIES_SP0[has_ht],
  dbh      = trees$DBH[has_ht],
  height   = trees$HT_PROJ[has_ht]
)

trees$MER_M3[has_ht] <- tree_volume(
  bec_zone    = trees$BEC_ZONE[has_ht],
  species     = trees$SPECIES_SP0[has_ht],
  dbh         = trees$DBH[has_ht],
  height      = trees$HT_PROJ[has_ht],
  volume_type = "MER"
)

vol_summary <- do.call(rbind, lapply(sp0_list, function(sp0) {
  sub <- trees[!is.na(trees$SPECIES_SP0) & trees$SPECIES_SP0 == sp0 &
               !is.na(trees$WSV_M3), ]
  data.frame(SP0      = sp0,
             n        = nrow(sub),
             WSV_mean = round(mean(sub$WSV_M3), 4),
             MER_mean = round(mean(sub$MER_M3), 4),
             MER_pct  = round(100 * mean(sub$MER_M3 / sub$WSV_M3), 1))
}))
print(vol_summary, row.names = FALSE)
```

The `MER_pct` column (merchantable fraction of WSV) reflects utilisation
efficiency; it tends to be lower for small-diameter trees where the
unmerchantable top represents a larger share of the stem.

---

---

## Function quick-reference

| Task | Function |
|------|----------|
| BEC-aware species standardisation | `species_correction(species, bec_zone)` |
| PSP code → SP0 (for volume **and** H-D model fitting) | `bc_species_to_sp0(corrected_code)` |
| PSP code → SP_TYPE conifer/deciduous (H-D fallback) | `bc_species_to_sp_type(corrected_code)` |
| PSP code → common name (for biomass) | `bc_species_to_biomass_name(species, bec_zone)` |
| H-D model fitting — all SP0 groups + SP_TYPE fallback | `fit_hd_models_by_group(measured)` |
| Height imputation with BLUP calibration | `ht_impute(trees, hd_result)` |
| H-D model fitting — single group (manual control) | `fit_hd_model(data, model, group_col, nested_col)` |
| Height prediction from DBH | `ht_from_dbh(dbh, model, a, b)` |
| Total aboveground biomass | `biomass_tree(species, dbh, height)` |
| Component biomass | `biomass_components(species, dbh, height)` |
| Whole-stem volume | `tree_volume(bec_zone, species, dbh, height)` |
| Merchantable volume | `tree_volume(..., volume_type = "MER")` |
| Log-section volume | `tree_volume_section(bec_zone, species, dbh, height, log_lengths)` |
