---
title: "Height-Diameter Modelling with PSP Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Height-Diameter Modelling with PSP Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

## Overview

Height-diameter (H-D) allometric models are a cornerstone of BC forest
inventory and growth modelling. In Permanent Sample Plot (PSP) programs,
tree heights are expensive to measure, so only a subset of trees in each
plot are typically measured for height. H-D models are fitted to those
measured trees and then used to **predict heights for all remaining
unmeasured stems**.

This vignette demonstrates the full workflow using simulated PSP data:

1. Simulate a multi-plot, multi-species PSP dataset with realistic missing heights  
2. Explore the observed H-D relationships  
3. Fit fixed-effects models and compare model forms  
4. Fit a mixed-effects model with a random plot effect  
5. Predict missing heights across the dataset  

---

## 1. Load PSP Data

The built-in `psp_trees` dataset contains 20 plots across four BEC zones with
five species. We use three species here (Douglas-fir `FDI`, Lodgepole pine
`PLI`, Western hemlock `HW`). Approximately 25% of trees have measured
heights; the remaining 75% are `NA`. The dataset also includes dead standing
trees (`LV_D == "D"`, ~11%) and broken-top trees (`BTOP == TRUE`, ~5%)
— both are excluded from H-D model training.

```{r load-data}
library(CanWestAllometryCFS)

sim_trees <- psp_trees[psp_trees$SPECIES %in% c("FDI", "PLI", "HW"), ]

# Species crosswalk — required for fit_hd_models_by_group() and ht_impute()
sim_trees$SPECIES_CORR    <- species_correction(sim_trees$SPECIES,
                                                sim_trees$BEC_ZONE)
sim_trees$SPECIES_SP0     <- bc_species_to_sp0(sim_trees$SPECIES_CORR)
sim_trees$SPECIES_SP_TYPE <- bc_species_to_sp_type(sim_trees$SPECIES_CORR)

cat(sprintf("Plots: %d   |   Trees: %d   |   Measured: %d   |   Missing: %d\n",
            length(unique(sim_trees$SITE_IDENTIFIER)),
            nrow(sim_trees),
            sum(!is.na(sim_trees$HEIGHT)),
            sum(is.na(sim_trees$HEIGHT))))
```

---

## 2. Explore the H-D Relationships

A quick look at the measured trees before fitting anything.

```{r explore}
# Training data: alive trees with measured height, no broken tops
measured <- sim_trees[!is.na(sim_trees$HEIGHT) &
                       sim_trees$LV_D == "L" &
                       !sim_trees$BTOP, ]

# Summary by species
sp_summary <- do.call(rbind, lapply(sort(unique(sim_trees$SPECIES)), function(sp) {
  sub <- measured[measured$SPECIES == sp, ]
  data.frame(Species = sp, N = nrow(sub),
             DBH_min = min(sub$DBH), DBH_max = max(sub$DBH),
             HT_min  = min(sub$HEIGHT), HT_max = max(sub$HEIGHT))
}))
print(sp_summary, row.names = FALSE)
```

```{r plot-hd, fig.cap = "Observed height-diameter relationships by species."}
sp_colours <- c(FDI = "#1b7837", PLI = "#762a83", HW = "#d6604d")

plot(HEIGHT ~ DBH, data = measured,
     col  = sp_colours[measured$SPECIES],
     pch  = 16, cex = 0.7,
     xlab = "DBH (cm)", ylab = "Height (m)",
     main = "Observed H-D relationships")
legend("topleft", legend = names(sp_colours),
       col = sp_colours, pch = 16, bty = "n")
```

---

## 3. H-D Model Fitting

`fit_hd_model()` fits a single H-D model to a set of measured trees.
It supports six model forms (`"naslund"`, `"curtis"`, `"korf"`,
`"weibull"`, `"richards"`, `"logistic"`); Naslund is the default and
generally performs well across BC species and size ranges.

Here we fit all forms to Douglas-fir (`FDI`) and rank by marginal R²:

```{r compare-models}
fd_data <- measured[measured$SPECIES == "FDI", ]

model_forms <- c("naslund", "curtis", "korf", "weibull", "richards")

results <- do.call(rbind, lapply(model_forms, function(m) {
  fit <- tryCatch(fit_hd_model(fd_data, model = m), error = function(e) NULL)
  data.frame(model   = m,
             r2_marg = if (is.null(fit)) NA else round(fit$r2_marg, 4))
}))

results <- results[order(-results$r2_marg, na.last = TRUE), ]
print(results, row.names = FALSE)
```

The differences between forms are often small when data cover a wide DBH
range. To inspect and visualise a specific fit:

```{r plot-fit, fig.cap = "Naslund model fitted to Fd trees (dashed curve)."}
fd_naslund <- fit_hd_model(fd_data, model = "naslund")
cat("Fitted coefficients:", round(fd_naslund$coefficients, 4), "\n")
cat("Marginal R²        :", round(fd_naslund$r2_marg, 4), "\n")

cf      <- fd_naslund$coefficients
dbh_seq <- seq(5, 65, by = 0.5)

plot(HEIGHT ~ DBH, data = fd_data,
     col  = "#1b7837", pch = 16, cex = 0.7,
     xlab = "DBH (cm)", ylab = "Height (m)",
     main = "FDI (Douglas-fir) \u2014 Naslund H-D model")
lines(dbh_seq,
      ht_from_dbh(dbh_seq, model = "naslund", a = cf["a"], b = cf["b"]),
      lwd = 2, lty = 2, col = "black")
legend("topleft", legend = c("Observed", "Fitted"), pch = c(16, NA),
       lty = c(NA, 2), col = c("#1b7837", "black"), bty = "n")
```

---

## 4. Mixed-Effects Model — Random Plot Effect

`fit_hd_model()` supports two levels of random effects via `group_col` and
`nested_col`:

- **`group_col`** (plot level) — a random effect on the asymptote parameter
  captures plot-to-plot productivity differences, improving predictions for
  trees on plots where some heights were measured.
- **`nested_col`** (tree level, nested within plot) — for multi-year PSP data
  where the same tree is measured repeatedly, adding a tree-level random
  effect accounts for individual tree-to-tree variation within a plot.
  Trees measured only once receive a tree-level BLUP of zero and fall back
  to the plot-level prediction.

For single-occasion data (each tree appears once), only `group_col` is
needed. For multi-year data, supply both.

```{r fit-mixed, eval = requireNamespace("nlme", quietly = TRUE)}
# Single-occasion: plot-level random effect only
fd_me <- fit_hd_model(fd_data, model = "naslund", group_col = "SITE_IDENTIFIER")

cat("Fixed coefficients :", round(fd_me$coefficients, 4), "\n")
cat("Marginal R²        :", round(fd_me$r2_marg, 4), "\n")
```

The marginal R² for the mixed-effects model is typically similar to or
slightly lower than the fixed-effects R², because the random effects absorb
some of the between-plot variance rather than the fixed-effects prediction.
The mixed-effects model is preferred when you have a clear grouping
structure and enough trees per group (generally ≥ 5–10).

---

## 5. Predict Missing Heights

For operational use across all species and plots, `fit_hd_models_by_group()`
and `ht_impute()` replace the manual per-species loop with a single call
each. `fit_hd_models_by_group()` handles the SP0 loop, SP_TYPE fallback,
and the mixed/fixed decision automatically. `ht_impute()` applies the
fitted models and BLUP calibration to fill all missing heights.

```{r predict-missing}
hd_result <- fit_hd_models_by_group(measured)
print(hd_result$summary, row.names = FALSE)

sim_trees <- ht_impute(sim_trees, hd_result, impute_btop = TRUE)

n_filled <- sum(is.na(sim_trees$HEIGHT) & !is.na(sim_trees$HT_PROJ))
cat(sprintf("%d of %d missing heights filled. Still missing: %d\n",
            n_filled,
            sum(is.na(sim_trees$HEIGHT)),
            sum(is.na(sim_trees$HT_PROJ))))
```

A sample of the predictions alongside the true (simulated) heights:

```{r prediction-check}
predicted_rows <- which(is.na(sim_trees$HEIGHT) & !is.na(sim_trees$HT_PROJ))
check <- sim_trees[head(predicted_rows, 10),
                   c("unitreeid", "SPECIES", "DBH", "HT_PROJ", "HEIGHT_TRUE")]
print(check, row.names = FALSE)
```

---

## Summary

| Step | Function |
|---|---|
| Fit a single model (one species/group) | `fit_hd_model(data, model)` |
| Add plot-level random effect | `fit_hd_model(data, model, group_col)` |
| Add tree-level random effect (multi-year) | `fit_hd_model(data, model, group_col, nested_col)` |
| Fit all SP0 groups + SP_TYPE fallback | `fit_hd_models_by_group(measured)` |
| Fill missing heights (operational) | `ht_impute(trees, hd_result)` |
| Predict from known coefficients | `ht_from_dbh()` / `ht_naslund()` etc. |
