---
title: "Tree Volume Estimation with PSP Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tree Volume Estimation with PSP Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

## Overview

Individual-tree gross volume is a core output of BC forest inventory and PSP
remeasurement programs. This vignette demonstrates how to compute tree volume
using the **Kozak (2002) KBEC variable-exponent taper equation** as implemented
in `tree_volume()`. The function accepts BEC zone and species codes that are
standard in BC PSP and inventory databases.

Topics covered:

1. Compute whole-stem (WSV), merchantable (MER), and stump volume for PSP trees
2. Understand the difference between volume types
3. Handle broken-top trees
4. Aggregate to plot-level growing stock
5. Bridge with H-D predictions when heights are missing

The taper equation requires both DBH and total height. See the
*Height-Diameter Modelling with PSP Data* vignette for filling missing heights
before running the volume calculations here.

---

## 1. Load PSP Data

The built-in `psp_trees` dataset contains 20 plots across four BEC zones
(SBS, ICH, CWH, IDF) with five BC species recorded as raw PSP/VRI inventory
codes (`AT`, `FDI`, `HW`, `PLI`, `SW`). The BEC-zone-aware crosswalk
(`species_correction()` → `bc_species_to_sp0()`) maps these to the SP0 group
codes required by `tree_volume()`. Volume requires a measured or predicted
height; here we use `HEIGHT_TRUE` (available for all trees) to focus on the
volume workflow. In practice, you would first predict missing heights using
the H-D modelling workflow (see `vignette("hd_psp_workflow")`).

`tree_volume()` looks up Kozak taper coefficients internally from
`taper_coefs_kbec` (KBEC) or `taper_coefs_kfiz3` (KFIZ3). Users never need
to access these tables directly, but they can be inspected to check which
species and zone combinations are supported:

```r
# Species/BEC combinations available for KBEC
unique(taper_coefs_kbec[, c("species", "bec_zone")])

# Species/FIZ combinations available for KFIZ3
unique(taper_coefs_kfiz3[, c("species", "fiz_zone")])
```

```{r load-data}
library(CanWestAllometryCFS)

# Use HEIGHT_TRUE so all trees have a height value for volume calculation.
# In practice, replace with predicted heights from fit_hd_model() first.
sim_trees <- psp_trees
sim_trees$HEIGHT      <- sim_trees$HEIGHT_TRUE
sim_trees$SPECIES_SP0 <- bc_species_to_sp0(
                           species_correction(sim_trees$SPECIES,
                                             sim_trees$BEC_ZONE))

cat(sprintf("Plots: %d   |   Trees: %d   |   BEC zones: %s\n",
            length(unique(sim_trees$SITE_IDENTIFIER)), nrow(sim_trees),
            paste(sort(unique(sim_trees$BEC_ZONE)), collapse = ", ")))
```

---

## 2. Whole-Stem Volume

`tree_volume()` accepts vectors and recycles scalar `bec_zone` or `species`
arguments, matching the conventions used in PSP data frames where one column
holds the BEC zone and another holds the species code.

```{r wsv}
sim_trees$WSV <- tree_volume(
  bec_zone = sim_trees$BEC_ZONE,
  species  = sim_trees$SPECIES_SP0,
  dbh      = sim_trees$DBH,
  height   = sim_trees$HEIGHT
)

cat(sprintf("Computed WSV for %d trees. NA: %d\n",
            nrow(sim_trees), sum(is.na(sim_trees$WSV))))

# Summary by species
sp_summary <- do.call(rbind, lapply(c("D","H","PL","S"), function(sp) {
  sub <- sim_trees[sim_trees$SPECIES_SP0 == sp & !is.na(sim_trees$WSV), ]
  if (nrow(sub) == 0) return(NULL)
  data.frame(
    species    = sp,
    n          = nrow(sub),
    dbh_mean   = round(mean(sub$DBH), 1),
    ht_mean    = round(mean(sub$HEIGHT), 1),
    wsv_mean   = round(mean(sub$WSV), 3),
    wsv_range  = paste0(round(range(sub$WSV), 3), collapse = "–")
  )
}))
print(sp_summary, row.names = FALSE)
```

```{r plot-wsv, fig.cap = "Whole-stem volume by DBH, coloured by species."}
sp_cols <- c(D = "#1b7837", H = "#762a83", PL = "#d6604d", S = "#4393c3")
vis     <- sim_trees[!is.na(sim_trees$WSV), ]

plot(WSV ~ DBH, data = vis,
     col  = sp_cols[vis$SPECIES_SP0],
     pch  = 16, cex = 0.7,
     xlab = "DBH (cm)", ylab = expression(WSV~(m^3)),
     main = "Whole-stem volume vs DBH")
legend("topleft",
       legend = names(sp_cols),
       col    = sp_cols,
       pch    = 16, bty = "n", title = "Species")
```

---

## 3. Volume Types: WSV, MER, and STUMP

Three volume types are available:

- **WSV** — whole-stem volume from ground to tip (stump cylinder +
  Smalian slices to the tip)
- **MER** — merchantable volume: WSV minus stump and minus the top section
  where inside-bark diameter falls below `utop_dib` (default 10 cm)
- **STUMP** — volume of the stump cylinder only

```{r volume-types}
# Compute all three types for a subset of trees
subset_idx <- which(sim_trees$DBH >= 20 & sim_trees$DBH <= 40)[1:8]
ex <- sim_trees[subset_idx, c("unitreeid", "SPECIES", "SPECIES_SP0", "DBH", "HEIGHT")]

ex$WSV   <- tree_volume(sim_trees$BEC_ZONE[subset_idx],
                        ex$SPECIES_SP0, ex$DBH, ex$HEIGHT,
                        volume_type = "WSV")
ex$MER   <- tree_volume(sim_trees$BEC_ZONE[subset_idx],
                        ex$SPECIES_SP0, ex$DBH, ex$HEIGHT,
                        volume_type = "MER")
ex$STUMP <- tree_volume(sim_trees$BEC_ZONE[subset_idx],
                        ex$SPECIES_SP0, ex$DBH, ex$HEIGHT,
                        volume_type = "STUMP")

ex[, c("WSV","MER","STUMP")] <- round(ex[, c("WSV","MER","STUMP")], 4)
print(ex, row.names = FALSE)
```

The MER-to-WSV ratio (`MER/WSV`) reflects utilisation efficiency and varies
by DBH: for small trees the top log is a larger proportion of the stem, so
the ratio tends to be lower.

```{r mer-ratio, fig.cap = "MER/WSV ratio as a function of DBH. Values below 1 indicate volume lost to stump and unmerchantable top."}
sim_trees$MER <- tree_volume(
  bec_zone    = sim_trees$BEC_ZONE,
  species     = sim_trees$SPECIES_SP0,
  dbh         = sim_trees$DBH,
  height      = sim_trees$HEIGHT,
  volume_type = "MER"
)

vis2 <- sim_trees[!is.na(sim_trees$WSV) & sim_trees$WSV > 0, ]
vis2$MER_RATIO <- vis2$MER / vis2$WSV

plot(MER_RATIO ~ DBH, data = vis2,
     col  = sp_cols[vis2$SPECIES_SP0],
     pch  = 16, cex = 0.7,
     xlab = "DBH (cm)", ylab = "MER / WSV",
     main = "Merchantable fraction of whole-stem volume",
     ylim = c(0, 1))
abline(h = 1, lty = 2, col = "grey50")
legend("bottomright",
       legend = names(sp_cols),
       col    = sp_cols,
       pch    = 16, bty = "n", title = "Species")
```

---

## 4. Broken-Top Trees

PSPs regularly contain broken-top trees. The `btop_height` parameter caps
the effective tree height used in the integration, so broken-top WSV is
computed up to the break point only.

```{r broken-top}
# One example tree across a range of break heights
ex_dbh  <- 35.0
ex_ht   <- 30.0
ex_bec  <- "CWH"
ex_sp   <- "H"

v_full  <- tree_volume(ex_bec, ex_sp, ex_dbh, ex_ht)
btops   <- seq(10, 28, by = 2)
# tree_volume iterates over trees; use sapply to vary btop_height for one tree
v_btop  <- sapply(btops, function(bt)
  tree_volume(ex_bec, ex_sp, ex_dbh, ex_ht, btop_height = bt))

cat(sprintf("Full WSV (30 m): %.4f m³\n", v_full))
print(data.frame(btop_height = btops, WSV = round(v_btop, 4)),
      row.names = FALSE)
```

```{r plot-btop, fig.cap = "Whole-stem volume loss as break height decreases below total height."}
plot(c(btops, ex_ht), c(v_btop, v_full),
     type = "b", pch = 16, col = "#762a83",
     xlab = "Break height (m)",
     ylab = "WSV (m^3)",
     main = sprintf("Broken-top volume  (DBH=%.0f cm, H=%.0f m, %s %s)",
                    ex_dbh, ex_ht, ex_bec, ex_sp))
abline(h = v_full, lty = 2, col = "grey50")
text(ex_ht - 1, v_full * 1.015, "full stem", cex = 0.75, col = "grey50")
```

---

## 5. Plot-Level Volume

Growing stock volume (m³/ha) is the most commonly reported PSP output.
Here we aggregate WSV to the plot level, assuming a 400 m² (0.04 ha) subplot.

```{r plot-volume}
PLOT_AREA_HA <- 0.04
EXP_FACTOR   <- 1 / PLOT_AREA_HA   # 25 trees-equivalent/ha

plot_vol <- do.call(rbind, lapply(unique(sim_trees$SITE_IDENTIFIER), function(pid) {
  sub   <- sim_trees[sim_trees$SITE_IDENTIFIER == pid & !is.na(sim_trees$WSV), ]
  data.frame(
    SITE_IDENTIFIER = pid,
    BEC_ZONE   = sub$BEC_ZONE[1],
    N_TREES    = nrow(sub),
    WSV_M3_HA  = round(sum(sub$WSV,   na.rm = TRUE) * EXP_FACTOR, 1),
    MER_M3_HA  = round(sum(sub$MER,   na.rm = TRUE) * EXP_FACTOR, 1)
  )
}))

print(plot_vol[order(plot_vol$WSV_M3_HA, decreasing = TRUE), ],
      row.names = FALSE)
cat(sprintf("\nMean WSV: %.1f m³/ha  |  Mean MER: %.1f m³/ha\n",
            mean(plot_vol$WSV_M3_HA),
            mean(plot_vol$MER_M3_HA)))
```

---

## 6. Bridging with H-D Predictions

In practice, PSP heights are partially measured. A typical pipeline:

1. Fit H-D models per species using `fit_hd_model()` (see H-D vignette)
2. Predict missing heights with `ht_from_dbh()`
3. Pass the complete height vector to `tree_volume()`

```{r hd-bridge}
# Introduce 60% missing heights to the dataset
set.seed(99)
sim_trees$HEIGHT_OBS <- sim_trees$HEIGHT
sim_trees$HEIGHT_OBS[sample(nrow(sim_trees), round(nrow(sim_trees) * 0.60))] <- NA

# Step 1: Fit per-species H-D models from measured trees
measured_trees <- sim_trees[!is.na(sim_trees$HEIGHT_OBS), ]
fits <- lapply(c("D", "H", "PL", "S"), function(sp) {
  sub <- measured_trees[measured_trees$SPECIES_SP0 == sp, ]
  if (nrow(sub) < 5) return(NULL)
  fit <- tryCatch(fit_hd_model(sub, model = "naslund"), error = function(e) NULL)
  list(sp = sp, fit = fit)
})

# Step 2: Predict missing heights
sim_trees$HT_PROJ <- sim_trees$HEIGHT_OBS
for (f in fits) {
  if (is.null(f$fit)) next
  cf      <- f$fit$coefficients
  missing <- is.na(sim_trees$HEIGHT_OBS) & sim_trees$SPECIES_SP0 == f$sp
  sim_trees$HT_PROJ[missing] <- round(
    ht_from_dbh(sim_trees$DBH[missing],
                model = "naslund", a = cf["a"], b = cf["b"]),
    1
  )
}

cat(sprintf("Heights filled: %d / %d  |  Still missing: %d\n",
            sum(!is.na(sim_trees$HT_PROJ)),
            nrow(sim_trees),
            sum(is.na(sim_trees$HT_PROJ))))

# Step 3: Compute volume on the complete height vector
complete <- !is.na(sim_trees$HT_PROJ)
sim_trees$WSV_FILLED <- NA_real_
sim_trees$WSV_FILLED[complete] <- tree_volume(
  bec_zone = sim_trees$BEC_ZONE[complete],
  species  = sim_trees$SPECIES_SP0[complete],
  dbh      = sim_trees$DBH[complete],
  height   = sim_trees$HT_PROJ[complete]
)

cat(sprintf("Volume computed for %d trees  (%.0f%% coverage)\n",
            sum(!is.na(sim_trees$WSV_FILLED)),
            100 * mean(!is.na(sim_trees$WSV_FILLED))))
```

---

## Summary

| Goal | Function / argument |
|------|---------------------|
| Whole-stem volume (m³) | `tree_volume(bec, sp, dbh, ht)` |
| Merchantable volume | `tree_volume(..., volume_type = "MER")` |
| Stump volume | `tree_volume(..., volume_type = "STUMP")` |
| Broken-top tree | `tree_volume(..., btop_height = h)` |
| Change stump height | `tree_volume(..., stump_height = 0.15)` |
| Change top DIB limit | `tree_volume(..., utop_dib = 7)` |
| Available species / BEC | `taper_coefs_kbec[, c("species","bec_zone")]` |
| View citations | `volume_citations()` |

```{r citations}
volume_citations()
```
